Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-15 manuscript.tex
ms edits

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated assuming no variance in female mating (e.g., $C^2_{f}=0$) and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Assuming the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}, Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio and variance in male reproductive success. Using these neutral predictions, we compared our empirically estimated diversity levels to normalized X/A and Y/A ratios of diversity. Note that we have only considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{Figure1.png}
\caption{The relation between relative effective population size and sex ratio bias for genes on autosomes (\textbf{A}) and sex chromosomes (\textbf{B}) (Y chromosome in red). The sex ratio is shown as the proportion of males, $N_{m}/(N_{f}+N_{m})$, where $N_{m}$ and $N_{f}$ are the effective number of breeding males and females, respectively, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$ and the $N_{e}$ for sex chromosomes and autosomes is given in equations 1-3. Dotted lines show predictions in the standard neutral model, where both males and females produce Poisson-distributed offspring numbers, where $N_{e{A}}=4N_{m}N_{f}/(N_{m}+N_{f})$ for autosomes, $N_{e{X}}=9N_{m}N_{f}/(4N_{m}+2N_{f})$ for the X, and $N_{e{Y}}=N_{m}/2$ for the Y \citep{wright1931evolution}. The solid curves correspond to increasing levels of variance in male reproductive success \citep{nomura2002effective} (see Methods). Assuming $\theta=4N_{e}\mu$ and equal neutral mutation rates among genes, the predicted $N_{e}$'s are used to generate null predictions for X/A and Y/A ratios of diversity.  
}%
\label{fig:spectrum}
\end{figure}

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by selection, we conducted two sets of forward-time simulations using the software SFSCODE \citep{hernandez2008flexible}: (i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection acting on beneficial mutations and reducing variation through selective sweeps (refs \X). For background selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ (range \X) from (\X) to \X (see SI text for simulation code).
To make the simulation output comparable to our data, we initialized simulations with our empirically estimated value of $\theta$ (from autosomes), and sampled 8 Y chromosomes per simulation from 24 populations. To consider the effects of selection against deleterious alleles (purifying selection), we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ from \X to \X. For simulating positive selection, we used \X [parameters here]. We ran 1000 replicates for each model and we calculated the average $\theta_{sim}$ to obtain the simulated Y/A ratio of neutral diversity under the both sets of models separately (purifying and positive selection). We tested the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but a Y-specific diversity reduction. Conceivably, low diversity on the Y could arise from a low Y-chromosome mutation rate or a lower mutation rate in males compared to females. However, recent evidence indicates that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, were not significantly different \citep{hough2014}.

% The pattern of reduced diversity on the Y chromosome is observed in both Africans and Europeans, suggesting that the effect is not population-specific, and holds regardless of whether the neutral sequence analyzed is near or far from genes (Table 1). Previous

%Diversity on the Y chromosome is likely not being under-estimated due to the inability to call variants in haploid regions of the genome because diversity on the X measured in females, where the X is diploid, is nearly identical to diversity on the X measured in males, where the X is haploid (Figure S2).

>>>>

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated using equations 1-3, assuming no variance in female mating (e.g., $C^2_{f}=0$), and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}, Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio and variance in male reproductive success. Using these neutral predictions, we compared our empirically estimated diversity levels to normalized X/A and Y/A ratios of diversity. Note that we have only considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{Figure1.png}
\caption{The relation between relative effective population size and sex ratio bias for genes on autosomes (\textbf{A}) and sex chromosomes (\textbf{B}) (Y chromosome in red). The sex ratio is shown as the proportion of males, $N_{m}/(N_{f}+N_{m})$, where $N_{m}$ and $N_{f}$ are the effective number of breeding males and females, respectively, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$ and the $N_{e}$ for sex chromosomes and autosomes is given in equations 1-3. Dotted curves show predictions in the standard neutral model, where both males and females produce Poisson-distributed offspring numbers and the chromosomal $N_{e}$'s are given by equations 4-6 \citep{wright1931evolution}. Solid curves correspond to increasing levels of variance in male reproductive success \citep{nomura2002effective} (see Methods). Assuming $\theta=4N_{e}\mu$ and equal neutral mutation rates among genes, the predicted $N_{e}$'s are used to generate null predictions for X/A and Y/A ratios of diversity.  
}%
\label{fig:spectrum}
\end{figure}

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by selection, we conducted two sets of forward-time simulations using the software SFSCODE \citep{hernandez2008flexible}: (i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection acting on beneficial mutations and reducing variation through selective sweeps (refs \X). For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ (range \X) from (\X) to \X and keeping the shape parameter, \X, constant (see Supporting Information for simulation code). To make the simulation output comparable to our data, we initialized simulations with our empirically estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations. For simulating positive selection, we used \X [parameters here]. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of models separately (purifying and positive selection). We then tested the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise at least in part from a low mutation rate on the Y, or a lower mutation rate in males compared to females. However, these possibilities can be rejected because there is no evidence that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are significantly different \citep{hough2014}. In addition, the pattern of reduced Y chromosome diversity occurs in each of the \textit{R. hastatulus} sub-clades that were inferred to be monophyletic (\X Figure 2; Table 1; Figure S1,S2), suggesting that the effect is not population-specific.
